### 2017-03-08 manuscript/manuscript.tex
discussion 70%

\usepackage{color, listings, bm, amsmath, blkarray, gensymb}
% \usepackage[percent]{overpic}
\usepackage{tikz}
% \usepackage{graphicx}
% \usepackage{epsfig}
% \usepackage{subfig}
% \captionsetup{belowskip=12pt}

\begin{document}
% todo:
% include snowmelt in model run. see if it's worth including with precip and temp as a climate predictor
% introduce terminology:
% responses = water temp, Q
% climatic predictors = air temp, precip, drought
% landscape predictors = many
%include hydrograph figures from kathy liederman and julian olden to show
    %rain dom, rain and snow, snow dom
%include the over-time plots in the supplemental information
%in regression figure, line up left side of legend box with plot above
    %make gray horiz line go only to topright corner of box
    %make axis lines and plot borders cleaner (right now axis segments are darker)
    %instead of elevation on x, use ice and color by elev (orange to blue) X no story
    %for loadings, color by ice and make it darkgreen , midgreen, white X not without the above
    % add % ice as one of the hidden trends 
%discharge regression combo plot is scrapped. no relationship with snowmelt?
%discharge only really relevant in the Q v temp plot.
%increase unit text size in fig 2
%change area over 1000m to % area over 1000m and replot
%change "dominated" to "dominant"
%change "fall" to "autumn"    
%dont size points in figure 2. include size in legend in fig 3

@@

\subsection*{Time series analysis}

Response time series were modeled using dynamic factor analysis (DFA; \colorbox{red}{\lstinline{Zuur et al. 2003}}), a multivariate technique that can be thought of as an analog to principal component analysis in the time domain. In DFA, response time series are fit with a linear combination of shared, random-walk trends (usually many fewer than the total number of response series), predictors (which can have unique effects on each response series), and random error. We chose DFA over a traditional multivariate state space approach for two reasons. First, it provides advantages in computational efficiency, as 1-5 shared trends often adequately capture variation across dozens of responses, and at much lower parameter cost. Second, in terms of identifying what drives the shared trends, having fewer of them allows for greater inferential parsimony. Being a multivariate technique, DFA also provides an advantage over univariate alternatives in that covariance structure among responses can be specified and compared. All models were fit using maximum likelihood estimation by automatic differentiation, with Template Model Builder software \colorbox{red}{\lstinline{Kristensen et al. 2015}}, which we called using package TMB in R \colorbox{red}{\lstinline{R Core team 2016...}}.

@@

After model selection, climatic predictor effect sizes for each river were back-transformed to their original scales and regressed against landscape predictors in order to identify possible watershed-scale controls on coupling. To achieve this, we amassed an additional dataset of landscape features. These were collected individually for each of the watersheds corresponding to our 24 river sites, using the EPA's StreamCat (stream-catchment) data library \colorbox{red}{\lstinline{cite}} and the National Hydrography Dataset (NHDPlusV2)\colorbox{red}{\lstinline{cite}}. Each site was mapped to an individual river reach, defined as a segment bounded on each end by a stream or river source, confluence, or mouth. The region contributing flow to this reach (its watershed) was then fetched, along with selected areal data, from the NHDPlusV2 database. Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), base flow index, soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), bedrock depth (cm), Base Flow Index (BFI; \%), runoff (mm mo\textsuperscript{-1}), percent perennial ice and snow (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011). We also calculated area over 1000 m (km\textsuperscript{2}) and mean slope (percent rise) by delineating watersheds from a digital elevation model in ArcMap \colorbox{red}{\lstinline{cite}}.

@@

Mean monthly temperature trends for the three river classes, aggregated across all 38 years of data, deviated by a minimum of 1.0\degree C in December, and a maximum of 3.9\degree C in July (Fig. 1). SD rivers remained approximately two degrees colder than their RS counterparts through mid-late summer, and 3-4 degrees colder than RD throughout spring and summer. RD rivers were consistently warmest throughout the year. In January, RS reached a minimum of 4.4\degree C, and did not significantly differ from SD \colorbox{red}{\lstinline{stats throughout this section}}. RD, meanwhile, only attained a minimum of 5.6\degree C. RS reached a peak summer temperature of 16.9\degree C in July, while RS and SD followed in August with peak temperatures of 15.5 and 13.5\degree C, respectively.

@@

% \begin{figure}[h!]
\begin{center}
% \includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}
\resizebox{\textwidth}{!}{\includegraphics{figures/12_air_water_discharge.pdf}}
% \caption{Epipolar Geometry}
% \label{fig:air_water_disch}
\end{center}
% \end{figure}
\textbf{Figure 1} Monthly mean T\textsubscript{water} by river class, and T\textsubscript{air} and Q across classes, from 1978 to 2015. All depicted series represent discrete data. \colorbox{red}{\lstinline{see appendix for standard deviations?}} \\[\baselineskip]
% The combined hydrograph of all rivers reveals two primary peaks, one beginning in late spring and the other extending from late fall to early winter, with a prominent trough in late summer. Spring peak discharge coincides noticeably with a separation in water temperature between SD and RS, while the summer trough coincides with separation of RD and T\textsubscript{air}. November marks both the autumn peak in discharge and the point at which T\textsubscript{air} falls below T\textsubscript{water}.

DFA results, aggregated across months and years for each site, revealed a trend toward reduced $T_{air}\rightarrow T_{water}$ coupling with greater watershed elevation ($p=0.04, R^2=0.18$; Fig. 2a). On average, a 1\degree C change in T\textsubscript{air} corresponded to a $0.53\pm 0.03$\degree C change in T\textsubscript{water} at RD, a $0.51\pm 0.08$\degree C change at RS, and a a $0.45\pm 0.17$\degree C change at SD sites. A similar trend was observed with respect to $Precip\rightarrow T_{water}$ coupling ($p=0.03, R^2=0.21$; Fig. 2b), where a monthly change in total precipitation of 1 cm corresponded to a $0.02\pm 0.009$\degree C change in T\textsubscript{water} for RD, $-0.003\pm 0.009$\degree C for RS, and $0.004\pm 0.02$\degree C for SD. There was no evidence of coupling between snowmelt and T\textsubscript{water} (Fig. 2c). It is of note that the strongest examples of $T_{air}\rightarrow T_{water}$ and $Precip\rightarrow T_{water}$ coupling were observed in the Duckabush River, while the weakest examples are from the Elwha River. Both rivers drain glaciers of the Olympic Mountain Range, and both are SD, though the Elwha's watershed is larger.

\begin{tikzpicture}
    \draw (0, 0) node[inner sep=0] {\includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}};
    \draw (-4.1, -1.5) node[text width=9.5em] {\normalsize \textbf{Figure 2} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}. Regression lines indicate slopes significant at $\alpha=0.05$.};
\end{tikzpicture}

\begin{center}
\resizebox{\textwidth}{!}{\includegraphics{figures/15_disch_vs_air_focal5.pdf}}
\textbf{Figure 1} chili chili
\end{center}

Body text

\section*{Discussion}
The relationship between river classification and factor loadings is less clear

>>>>

\usepackage{color, listings, bm, amsmath, blkarray, gensymb}
% \usepackage[percent]{overpic}
\usepackage{tikz}
% \usepackage{graphicx}
% \usepackage{epsfig}
% \usepackage{subfig}
% \captionsetup{belowskip=12pt}
% \hfuzz=5.002pt  %suppress overfull warnings

\begin{document}
% todo:
% include snowmelt in model run. see if it's worth including with precip and temp as a climate predictor
% introduce terminology:
% responses = water temp, Q
% climatic predictors = air temp, precip, drought
% landscape predictors = many
%include hydrograph figures from kathy liederman and julian olden to show
    %rain dom, rain and snow, snow dom
%include the over-time plots in the supplemental information
%in regression figure, line up left side of legend box with plot above
    %make gray horiz line go only to topright corner of box
    %make axis lines and plot borders cleaner (right now axis segments are darker)
    %instead of elevation on x, use ice and color by elev (orange to blue) X no story
    %for loadings, color by ice and make it darkgreen , midgreen, white X not without the above
    % add % ice as one of the hidden trends
%discharge regression combo plot is scrapped. no relationship with snowmelt?
%discharge only really relevant in the Q v temp plot.
%increase unit text size in fig 2
%change area over 1000m to % area over 1000m and replot
%change "dominated" to "dominant"
%change "fall" to "autumn"
%dont size points in figure 2. include size in legend in fig 3
%adjust letter position in corners of fig 2 (anchoring)
%see if its possible to directly compare Z and D
%include the average effect size of tair on twater in 3a
% can i say how much elevation would inflence temperature independent of snowmelt?
%examine change over time in just the elwha

@@

\subsection*{Time series analysis}
Response time series were modeled using dynamic factor analysis (DFA; \colorbox{red}{\lstinline{Zuur et al. 2003}}), a multivariate technique that can be thought of as an analog to principal component analysis in the time domain. In DFA, response time series are fit with a linear combination of shared, random-walk trends (usually many fewer than the total number of response series), predictors (which can have unique effects on each response series), and random error. We chose DFA over a traditional multivariate state space approach for two reasons. First, it provides advantages in computational efficiency, as 1-5 shared trends often adequately capture variation across dozens of responses, and at much lower parameter cost. Second, in terms of identifying what drives the shared trends, having fewer of them allows for greater inferential parsimony. Being a multivariate technique, DFA also provides an advantage over univariate alternatives in that covariance structure among responses can be specified and compared. All models were fit using maximum likelihood estimation by automatic differentiation, with Template Model Builder software \colorbox{red}{\lstinline{Kristensen et al. 2015}}, which we called using package TMB in R \colorbox{red}{\lstinline{R Core team 2016...}}.

@@

After model selection, climatic predictor effect sizes for each river were back-transformed to their original scales and regressed against landscape predictors in order to identify possible watershed-scale controls on coupling. To achieve this, we amassed an additional dataset of landscape features. These were collected individually for each of the watersheds corresponding to our 24 river sites, using the EPA's StreamCat (stream-catchment) data library \colorbox{red}{\lstinline{cite}} and the National Hydrography Dataset (NHDPlusV2)\colorbox{red}{\lstinline{cite}}. Each site was mapped to an individual river reach, defined as a segment bounded on each end by a stream or river source, confluence, or mouth. The region contributing flow to this reach (its watershed) was then fetched, along with selected areal data, from the NHDPlusV2 database. Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), base flow index, soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), bedrock depth (cm), Base Flow Index (BFI; \%), runoff (mm mo\textsuperscript{-1}), percent perennial ice and snow (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011). Monitoring site elevation was also included. Finally, we calculated area over 1000 m (km\textsuperscript{2}) and mean slope (percent rise) by delineating watersheds from a digital elevation model in ArcMap \colorbox{red}{\lstinline{cite}}.

@@

Mean monthly temperature trends for the three river classes, aggregated across all 38 years of data, deviated by a minimum of 1.0\degree C in December, and a maximum of 3.9\degree C in July (Fig. 2). SD rivers remained approximately two degrees colder than their RS counterparts through mid-late summer, and 3-4 degrees colder than RD throughout spring and summer. RD rivers were consistently warmest throughout the year. In January, RS reached a minimum of 4.4\degree C, and did not significantly differ from SD \colorbox{red}{\lstinline{stats throughout this section}}. RD, meanwhile, only attained a minimum of 5.6\degree C. RS reached a peak summer temperature of 16.9\degree C in July, while RS and SD followed in August with peak temperatures of 15.5 and 13.5\degree C, respectively.

@@

% \begin{figure}[h!]
\begin{center}
% \includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}
\resizebox{\textwidth}{!}{\includegraphics{figures/12_air_water_discharge.pdf}}
% \caption{Epipolar Geometry}
% \label{fig:air_water_disch}
\end{center}
% \end{figure}

\textbf{Figure 2} Monthly mean T\textsubscript{water} by river class, and T\textsubscript{air} and Q across classes, from 1978 to 2015. All depicted series represent discrete data. \colorbox{red}{\lstinline{see appendix for standard deviations?}} \setlength{\parskip}{6pt}\\[\baselineskip]

The combined hydrograph of all rivers reveals two primary peaks, one beginning in late spring and the other extending from late fall to early winter, with a prominent trough in late summer. Spring peak discharge coincides noticeably with a separation in water temperature between SD and RS, while the summer trough coincides with separation of RD and T\textsubscript{air}. November marks both the autumn peak in discharge and the point at which T\textsubscript{air} falls below T\textsubscript{water}.

\sloppy
DFA results, aggregated across months and years for each site, revealed a trend toward reduced $T_{air}\rightarrow T_{water}$ coupling with greater watershed elevation ($p=0.04, R^2=0.18$; Fig. 3a). On average, a 1\degree C change in T\textsubscript{air} corresponded to a $0.53\pm 0.03$\degree C change in T\textsubscript{water} at RD, a $0.51\pm 0.08$\degree C change at RS, and a $0.45\pm 0.17$\degree C change at SD sites. A similar trend was observed with respect to $Precip\rightarrow T_{water}$ coupling ($p=0.03, R^2=0.21$; Fig. 3b), where a monthly change in total precipitation of 1 cm corresponded to a $0.02\pm 0.009$\degree C change in T\textsubscript{water} for RD, $-0.003\pm 0.009$\degree C for RS, and $0.004\pm 0.02$\degree C for SD. There was no evidence of coupling between snowmelt and T\textsubscript{water} (Fig. 3c), but this predictor was included in the most parsimonious DFA model selected via AIC and R\textsuperscript{2} \colorbox{red}{\lstinline{see appendix}}. It is of note that the strongest examples of $T_{air}\rightarrow T_{water}$ and $Precip\rightarrow T_{water}$ coupling were observed in the Duckabush River, while the weakest examples are from the Elwha River. Both rivers drain glaciers of the Olympic Mountain Range, and both are SD, though the Elwha's watershed is larger.

\begin{tikzpicture}
    \draw (0, 0) node[inner sep=0] {\includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}};
    \draw (-4.1, -1.5) node[text width=9.5em] {\normalsize \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}. Regression lines indicate slopes significant at $\alpha=0.05$.};
\end{tikzpicture}

In addition to the three climate predictors above, the best T\textsubscript{water} model also included five shared trends. Of these, four correlated significantly with at least one known watershed predictor. Figure 3 depicts the strongest correlated variables with each trend (insets d-e). These are, in arbitrary order of relevance, mean water table depth ($p<0.001, R^2=0.60$; Fig. 3d), \% glaciation ($p<0.01, R^2=0.30$; Fig. 3e), BFI ($p=0.01, R^2=0.25$; Fig. 3f), and mean slope ($p<0.01, R^2=0.29$; Fig. 3g). The fifth shared trend was not correlated with any variables in the watershed predictor dataset.

\begin{center}
\resizebox{\textwidth}{!}{\includegraphics{figures/15_disch_vs_air_focal5.pdf}}
\textbf{Figure 4} Relationship between $T_{air}\rightarrow T_{water}$ and $T_{air}\rightarrow Q$. Both axes are expressed per 1\degree C change in $T_{air}$. The red quadrant designates proportionality between all three variables, the blue inverse proportionality between each response and $T_{air}$. Regression lines indicate slopes significant at $\alpha=0.05$.
\end{center}

To examine possible sub-season interactions between T\textsubscript{air}, T\textsubscript{water} and Q, we performed an additional DFA with Q as the response. In both models, T\textsubscript{air} was allowed to have unique monthly effects. These effects, taken together, can be understood in relation to the four quadrants of the Cartesian coordinate system (increasing clockwise from upper right; Fig. 4).

In mid-winter (exemplified by February), all river classes primarily occupy the first quadrant, signifying $T\textsubscript{air} \propto T\textsubscript{water}$ and $T\textsubscript{air} \propto Q$, where $\propto$ denotes proportionality. RD shows the weakest Q response. By spring, many RS and SD sites develop an inverse relationship between T\textsubscript{air} and T\textsubscript{water}, denoted $T\textsubscript{air} \propto \frac{1}{T\textsubscript{water}}$, while RD sites change little from their winter state. June and August see a procession of most sites into the near fourth quadrant, with SD trailing. This signifies $T\textsubscript{air} \propto \frac{1}{Q}$, though $T\textsubscript{air} \propto T\textsubscript{water}$ remains. One stark exception is again the Elwha river, which occupies quadrant three. By fall, RS and SD have begun progress back toward their winter states, led by SD. RD, meanwhile, remain essentially unmoved from summer.

\section*{Discussion}

The effects of climate on T\textsubscript{water}, determined by dynamic factor analysis, suggest that nearly all rivers included in our dataset were influenced strongly by air temperature, precipitation, and/or snowmelt across 38 years of monthly data (Fig. 3). At most monitoring sites, T\textsubscript{water} closely tracked changes in T\textsubscript{air}, on average responding to increases and decreases with proportional movements of up to 66\% magnitude. However, some rivers only weakly track T\textsubscript{air}, and several patterns in the intensity of this coupling correlate strongly with watershed features relating to ice, groundwater, and slope. Glaciation and yearly snow burden are prominent among these, and for reasons of ecological and hydrological implication, the primary focus of the following discussion.

Without any analysis, a "buffering" effect (hereafter contrasted with "coupling") of ice on river temperature can be seen in the yearly patterns of \textsubscript{water} relative to T\textsubscript{air} (Fig. 2). The aggregate hydrograph peaks due to snowmelt from April to June, at the same time that the trajectories of RS and SD (snow-influenced rivers) start to drop off relative to RD. After snowmelt begins to subside, RS and SD recover with a noticeable jump. For rivers that receive glacial runoff (SD), this effect appears to remain, buffering them from summer temperature rise where RS rivers instead take on the character of RD (Fig. 4). In an extreme case, the Elwha River was actually cooler in August during those years in which air temperature was higher, likely due to increased runoff from Carrie and Eel glaciers. The buffering effect of ice on river temperature is therefore two-fold, acting first on all snowmelt-influenced rivers through a cold-water pulse in spring, and then on a subset of those rivers throughout summer and fall, by way of glacial runoff. For RD rivers, which receive little to no input from ice, summer temperature is entirely dictated by that of the surrounding air, and whatever rain falls through it. Though higher-elevation watersheds will always produce colder water, independent of the influence of ice, it can be expected that RS and SD rivers will grow more similar to RD as regional temperatures warm and glaciers decline. That is to say, formerly reliably cold-water streams and associated habitats may see increases in both summer and winter average temperatures, as well as higher variability from year to year. The Elwha in particular may slip from its current state of high resistance to seasonal climatic changes. We tested for changes in mean and variance of $T\textsubscript{air}\rightarrow T\textsubscript{water}$ and $T\textsubscript{air}\rightarrow Q$ coupling between 1978 and 2015, but did not detect any widespread patterns \colorbox{red}{\lstinline{appendix B}}.

In addition to the three climate predictors, five shared trends were fit by the most parsimonious DFA model. These represent additional drivers responsible for structuring water temperature across some or all of the 24 sites included in the analysis. Each monitoring sites' factor loading on a particular shared trend indicates the degree to which the trend accounted for variance in T\textsubscript{water} at that site. While the precise identities of these drivers cannot be attained with certainty, they can be inferred through correlation with predictor variables. In this way, we determined the most likely landscape drivers of T\textsubscript{water} to be perennial ice and snow cover, mean watershed slope, and groundwater influx. Slope's likely mechanism of influence is increased turbulence and mixing of water and air in steep, headwater streams, which allows convective warming and cooling to occur more rapidly \colorbox{red}{\lstinline{cite}}.

%experiment with vim marks

% single out elwha as mostly unaffected by climate
% explain that the effect has not changed over time (nor the variance)

% fig1: buffering effect is visible as damping followed by jump in RS SD
    SD RS colder throughout the year, less influenced by changes in air.
    % spring snowmelt (hydrograph peak) is the most significant item
    RD actually remains warmer than air throughout the spring, and matches it through July.
% fig2: snow/glac streams are buffered, overall, from temp changes due to air and precip.
    snow isn't signif because of the arrangement of points along the x axis. however, there's a negative
        snowmelt->Twater for the glac streams and a positive one for the RS streams
    Why might each of the landscape predictors be such?
        water table depth: ground water farther from surface, less responsive ground water (to whatever it is)
        BFI: also ground water influx
        ice snow: separation of RD RS?: this could be the rain-on-snow effect
        slope: faster and more turbulent water, greater mixing; also correlated with snow/glac
    what could the other factor be (trend 3)? random error, or a geological or morphological or microclimatic feature not captured by our watershed predictors 
fog3: step through

The relationship between river classification and factor loadings is less clear

\section*{Appendix A}

DFA plots here

\section*{Appendix B}

over-time plots here
